Legacy Data Confounds Genomics Studies

### 2018-10-23 legacy.tex
updated text, added figures for indels and repeats

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Genomics Studies}
%\title{Batch Effects in the 1000 Genomes Project Data Confound Modern Genomics Studies}

@@

@@

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs and 1134 statistically significant insertions and deletions (indels) (Figure \ref{Manhattan} and Figure \ref{indels_Manhattan}). \sgcomment{We should discuss significance threshold and FDR!}

\begin{figure*}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}

\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
The line plot underneath shows the number of populations for which a variant has reached significance.
The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
The same variants identified as being low quality independently in each population are found in other populations. }
  \label{OverLap}
\end{figure*}

@@

Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 3990 SNPs and 1134 indels were found to be statistically significance. See supplementary file {here}. \luke{make this file}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_adjusted.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{Manhattan}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_indels.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 1134 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{indels_Manhattan}
\end{figure}

@@

@@

%\begin{center}
 %\begin{tabular}{l l l r r} 
 %\hline
%Author & Year & Journal & RSID & PMID \\ [0.5ex] 
 %\hline
%Ebejer JL	& 2013	 & Twin Res Hum Genet & 	rs6057648&23527680\\

%Mandage R &	2017	 & PLoS One	 & rs200655768, rs184202621,&28654678\\
%& & & rs201255786, rs201761909&\\
%\hline
%\end{tabular}
%\end{center}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Genomics Studies}

@@

The GWAS approach similarly identified $Q$-associated SNPs in 24 of the 26 populations in the 1kGP, with the phase 1 populations being most affected.
Over 3517 variants were independently associated to low $Q$ in at least two populations  (Figure \ref{OverLap}). 
To build a test statistic to represent the association across all populations, we summed the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified a total of 7971 statistically significant variants (Table \ref{sigTable}, Figure \ref{NRS_Manhattan}, Figure \ref{NRI_Manhattan}, Figure \ref{RS_Manhattan} and Figure \ref{RI_Manhattan}).
To account for the large number of tests we performed a two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the p-values using a nominal Type-I error rate $\alpha = 0.01$ \citep{Benjamini2006}.
\sgcomment{We should discuss significance threshold and FDR!}
\luke{is this more appropriate for the methods section?}

\begin{table}[h]
\begin{tabular}{l  l  l}
                      & \textbf{Repeat}  & \textbf{Non-Repeat}       \\ \cline{2-3} 
\textbf{SNP}  & 2207 & 3839 \\ \cline{2-3} 
\textbf{INDEL} & 750  & 1175 \\ \cline{2-3}
\end{tabular}
\caption{Number of statistically significant variants per category}
\label{sigTable}
\end{table}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{./Figures/SNPOverlap6.jpg}

\caption{Overlap of SNPs identified independently to be associated with average quality of mapped bases $Q$. 
The size of the crosses ( + ) are proportional to the -log10(p) value of that SNP.
The x axis is ranked by the frequency of occurrence of a SNP, then by genomic position.
The line plot underneath shows the number of populations for which a variant has reached significance.
The populations that tend to have the most individuals with low $Q$ also tend to have the most variants associated to $Q$. 
The same variants identified as being low quality independently in each population are found in other populations. }
  \label{OverLap}
\end{figure}

@@

Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 7971 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_SNPs.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 3839 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
 \label{NRS_Manhattan}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_NORepeat_Indels.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 1175 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
 \label{NRI_Manhattan}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_SNPs.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 2207 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
\label{RS_Manhattan}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/ManhattanPlot_Repeat_Indels.jpg}

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 750 insertions and deletions that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
 \label{RI_Manhattan}
\end{figure}

@@

\begin{figure}[h]
\includegraphics[width=\hsize,keepaspectratio]{./Figures/RegressionPlot_diabetes.jpg}
\caption{Logistic regression of rs301 to the average quality per mapped bases $Q$.}
\label{DiabetesSNP}
\end{figure}

@@
